---
title: "Analyses master thesis: David Crameri"
date: "07/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load data
source("R/load_data.R")
source("R/mytheme.R")

#packages
library(stringr)
library(ggplot2)
library(wesanderson)

```


```{r check out data, include=FALSE}

#check data structure
glimpse(data)  


#fehlende_zahne_fdi
sort(unique(data$fehlender_zahn_fdi_quadrant_zahnzahl))


```



### sample description (n = 622)

```{r sample description}

# plot gender
data %>% 
  filter(!duplicated(ID)) %>% 
  group_by(gschlecht_m_w) %>% 
  summarise(tot = n()) %>% 
  mutate(pct = tot / sum(tot)) %>% 
  ungroup() %>% 
  mutate(gender = if_else(gschlecht_m_w == "m", "Männlich", "Weiblich")) %>% 
  mutate(xlab_1 = paste0("(n = ", tot, ")"),
         xlab = paste(gender, xlab_1, sep = "\n")) -> dp


#print df
print(dp)


#plot
p <- ggplot(data = dp, 
       aes(x = xlab, y = pct, fill = gender)) +
  geom_bar(stat = "identity", width = .6) +
  labs(x = "", y = "Geschlecht in Prozent") + # does dot sound right?
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = wes_palette("Royal2")) +
  guides(fill = "none") + # check for further info: https://ggplot2.tidyverse.org/reference/guide_legend.html
  geom_text(label = scales::percent(dp$pct, accuracy = 1), size = 8, position = position_stack(vjust = .5)) +
  mytheme

print(p)

#save plot
ggsave(filename = here::here("output/plots/202207_overall_sample_gender.pdf"),
       plot = p,
       width = 6,
       height = 6,
       dpi = 300,
       device = cairo_pdf
)


#age at operation
data %>% 
  filter(!duplicated(ID)) -> dp
psych::describe(dp$alter_beim_opt)

#unique persons in sample
sum(!duplicated(data$ID))

```

### FF1: final sample description (n = 563)

```{r}

#1. gender
# extractet cases were eliminated (n = 9): #is that correct @david?
data %>% 
  #no duplicates
  filter(!duplicated(ID)) %>%
  #age older than 9
  filter(alter_beim_opt >= 9) %>% 
  # etrahierte zähne = nein
  filter(is.na(extrahierte_zahne)) -> dp

#filter gender overall
dp %>% 
  group_by(gschlecht_m_w) %>% 
  summarise(tot = n()) %>% 
  mutate(pct = tot / sum(tot)) %>% 
  ungroup() %>% 
  mutate(gender = if_else(gschlecht_m_w == "m", "Männlich", "Weiblich")) %>% 
  mutate(xlab_1 = paste0("(n = ", tot, ")"),
         xlab = paste(gender, xlab_1, sep = "\n")) -> dp_s

print(dp_s)

#plot
pq <- ggplot(data = dp_s, 
       aes(x = xlab, y = pct, fill = gschlecht_m_w)) +
  geom_bar(stat = "identity", width = .6) +
  labs(x = "", y = "Geschlecht in Prozent") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(label = scales::percent(dp_s$pct, accuracy = 1), size = 8, position = position_stack(vjust = .5)) +
  guides(fill = "none") +
  scale_fill_manual(values = wes_palette("Royal2")) +
  mytheme

print(pq)

#save:
ggplot2::ggsave(filename = "output/plots/202207_gender_final_subset.pdf",
       plot = pq,
       width = 6,
       height = 6,
       dpi = 300,
       device = cairo_pdf
)


#2. gender and age: MEAN
dp %>% 
  group_by(gschlecht_m_w) %>% 
  summarise(mean = mean(alter_beim_opt, na.rm = TRUE), 
            sd = sd(alter_beim_opt), #standard deviation
            se = sd(alter_beim_opt)/sqrt(length(alter_beim_opt))) #standard error of the mean

#3. missing teeth overall (n = 39)
dp %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl != 0) -> dp_m

head(dp_m,2)

#export data
write.csv(dp_m, here::here("output/datasets//202207_clients_hypodontia_overall.csv"))



```


### FF2: Nichtanlage vs. Gesund 

```{r}

#nach geschlecht
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(!duplicated(.$ID)) -> dp_d #duplicated cases are excluded

#summaries per gender and nichtanlage
dp_d %>% 
  group_by(gschlecht_m_w, nichtanlage_ja_nein) %>% 
  summarise(tot = n()) %>% 
  mutate(pct = tot / sum(tot))  %>%
  ungroup() -> df


#Add text to xlab
txt <- df %>% 
  group_by(gschlecht_m_w) %>% 
  summarise(gender_tot = sum(tot)) %>% 
  mutate(gender = if_else(gschlecht_m_w == "m", "Männlich", "Weiblich")) %>% 
  mutate(xlab_1 = paste0("(n = ", gender_tot, ")"),
         xlab = paste(gender, xlab_1, sep = "\n"))

#merge back
df_plot <- df %>% 
  left_join(., txt, by = "gschlecht_m_w")
  
print(df)

#plot
p <- ggplot(df_plot, aes(x = xlab, y = pct, fill = nichtanlage_ja_nein)) +
  geom_bar(stat = "identity", position = "stack") +
  labs( x = "", y = "Anzahl Klienten in Prozent") + #title ok like this?
  guides(fill = guide_legend(title = "Nichtanlage")) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = wes_palette("Royal2")) +
  geom_text(label = scales::percent(df_plot$pct, accuracy = 1), position = position_stack(vjust = .5), size = 8) +
  mytheme

print(p)

#save
ggplot2::ggsave(filename = "output/plots/202207_gender_nichtanlage.pdf",
        plot = p,
        width = 7,
        height = 6,
        device = cairo_pdf)

#do that correlation differ from a random result
#from results first into a matrix
xtabs(~ nichtanlage_ja_nein + gschlecht_m_w, data = as.data.frame(dp_d)) -> df_chi #which datas et should be used: with duplicated number or not

chi_sq <- chisq.test(df_chi) # both data sets does not show any assimilation between gender and nichtanlage
chi_sq

```


### FF3 & FF4: missing teeth description

> attention: missing teeth do not correspond with the sample size!

```{r}
#missing single tooth overall
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) -> dp_d  #exclude all persons without missing teeth
  

#missing teeth per person
dp_d %>% 
  group_by(ID) %>% 
  summarise(teeth_miss = n()) %>% 
  ungroup() -> df


#aggregate data set from above on idividual level: overall----
dp_ind <- df %>% 
  group_by(teeth_miss) %>% 
  summarise(tot_persons = n()) %>% 
  ungroup() %>% 
  mutate(pct = tot_persons / sum(tot_persons))

#attention here are all cases again summed up
#in other words 39 person have 56 missing teeth
print(dp_ind)

write.csv(file = "output/datasets/202207_table_missing_teeth_overall.csv", dp_ind)


#missing teeth per person: meaning most people misses 1 and 2 teeth
p <- ggplot(dp_ind, aes(x = as.factor(teeth_miss), y = pct, fill = as.factor(teeth_miss))) +
  #ev. as stacked bars? however data needs to be re-adjusted
  geom_bar(stat = "identity", position = position_dodge(), width = .6) +
  geom_text(label = dp_ind$tot_persons, position = position_stack(vjust = .5), size = 8) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = wes_palette("Royal1")) +
  labs(y = "Anteil fehlende Zähne in Prozent", x = "Anzahl fehlende Zähne") +
  guides(fill = "none") +
  mytheme

print(p)

#save plot
ggplot2::ggsave(filename = "output/plots/202207_missing_teeth.pdf",
                device = cairo_pdf,
                width = 6,
                height = 6,
                plot = p)


#another table: number of missing tooth buds
dp_d %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>% 
  group_by(fehlender_zahn_fdi_quadrant_zahnzahl) %>% 
  summarise(tot = n()) %>% 
  #calculate percentage
  mutate(pct_number_missing = tot / sum(tot) * 100) %>% 
  ungroup() %>% 
  #divide by number of persons in data set: 39
  mutate(pct_ind_hypodontia = tot / 39 * 100) -> df_e

#safe
write.csv(file = "output/datasets/202207_table_number_missing_tooth.csv", df_e)
    
    
```


### FF5: distribution of missing teeth
```{r, include=TRUE}
#overall, attention some persons are double
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>% 
  # create new variables
  mutate(quadrant = stringr::str_sub(fehlender_zahn_fdi_quadrant_zahnzahl, end = 1),
         position = stringr::str_sub(fehlender_zahn_fdi_quadrant_zahnzahl, start = 2)) %>%
  group_by(quadrant, position) %>% 
  summarise(tot = n()) %>% 
  mutate(pct = tot / sum(tot)) %>% 
  ungroup() -> dp

print(dp)

#xlab
txt <- dp %>% 
  group_by(quadrant) %>% 
  summarise(tot_q = sum(tot)) %>% 
  ungroup() %>% 
  mutate(xlab1 = paste0("n = (", tot_q, ")"))


#merge back data
df_plot <- dp %>% 
  left_join(., txt, by = "quadrant") %>% 
  mutate(xlab = paste(quadrant, xlab1, sep = "\n"))
  
#plot
p <- ggplot(df_plot, aes(x = xlab, y = pct, fill = position)) +
  geom_bar(stat = "identity", width = .6) +
  scale_y_origin(label = scales::percent) +
  guides(fill = guide_legend(title = "Position \nfehlende Zähne")) +
  geom_text(label = scales::percent(dp$pct, accuracy = 1), position = position_stack(vjust = .5), size = 8) +
  scale_fill_manual(values = wes_palette("Royal1")) +
  labs(x = "Quadrant", y = "Anteil fehlende Zähne in Prozent") +
  mytheme


print(p)

#save plot
ggplot2::ggsave(filename = "output/plots/202207_missing_teeth_per_quadrants.pdf",
                device = cairo_pdf,
                width = 9,
                height = 8,
                plot = p)

```



### some additional tables

> um keine verzerrungen zu bekommen müsste man mit 56 cases und nicht 39 personen arbeiten!
> hab nun die 56 als gesamtheit genommen (anstelle die 39 personen)

```{r}

#TODO: @gianna: kiefer und seite jeweils auf Geschlechter unterschiede überprüfen => chi-square test

#overall, attention some persons are double!
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>%
  mutate(quadrant = stringr::str_sub(fehlender_zahn_fdi_quadrant_zahnzahl, 
                                     end = 1)) %>% 
  mutate(quadrant = as.numeric(quadrant)) %>%
  select(ID,gschlecht_m_w, quadrant, fehlender_zahn_fdi_quadrant_zahnzahl)-> df_t



#table gender and chi-square
df_t %>% 
  group_by(gschlecht_m_w) %>% 
  summarise(tot_g = n()) %>% 
  mutate(pct = tot_g / sum(tot_g)) -> df

print(df)

write.csv(file = "output/datasets/202207_table_gender_hypodontia.csv", df)

#binomial test!
binom.test(15, 39) #gender and prevalance of hypodontia do not differ statistically significant
binom.test(21, 56) #wibliche patienten haben eine höhere prevalence von hypodonie

#further info about binomial test: https://www.statology.org/binomial-test-r/


#table unter/oberkiefer--------
df_t %>% 
  group_by(ID) %>% 
  summarise(tot = sum(quadrant)) %>% 
  ungroup() %>%
  right_join(., df_t[, c("ID", "quadrant", "gschlecht_m_w")], by = "ID")-> tt

# add new variable "kiefer"
tt %>% 
  # mutate(tot = if_else(tot == 3, 2.9, tot)) %>% 
  #attention order plays an important role
  mutate(kiefer = case_when((tot == 5 | tot == 9 | tot == 10) ~ "beide",
                            (quadrant == 1 | quadrant == 2) ~ "oben",
                            (quadrant == 3 | quadrant == 4 | tot == 3) ~ "unten",
                            TRUE ~ "beide")) ->  df_kiefer 

#one case needs to be change by hand (because sum 1, 3 is 4)
df_kiefer[df_kiefer$ID == 472,]$kiefer <- "beide"


#1. group on teeth level
df_kiefer %>% 
  group_by(kiefer, gschlecht_m_w) %>% 
  summarise(total = n()) %>% 
  mutate(pct = total / sum(total)) -> df_teeth

print(df_teeth)

# TODO David: Zahlen überprüfen
write.csv(file = "output/datasets/202207_table_kiefer_teeth_hypodontia.csv", df_teeth)


#double check it
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>%
  #change some cases: o, o to o; u,u to u, o, u to beide
  mutate(oben_unten = str_replace(oben_unten, "o, o", "o")) %>%
  mutate(oben_unten = str_replace(oben_unten, "u, u", "u")) %>% 
  mutate(oben_unten = str_replace(oben_unten, "o, u", "beide")) %>%
  # filter(!is.na(oben_unten)) %>% 
  group_by(oben_unten, gschlecht_m_w) %>% 
  summarise(total = n()) %>%
  mutate(pct = total / sum(total))-> df_test

#prepare data for chi_sq
xtabs(~ kiefer + gschlecht_m_w, data = df_kiefer) -> df_chisq

#chisquare test 
chisq.test(df_chisq) # no significance

#2. group on a individual level
df_kiefer %>% 
  #drop dups
  filter(!duplicated(ID)) %>%
  group_by(kiefer, gschlecht_m_w) %>% 
  summarise(total = n()) %>% 
  mutate(pct = total / sum(total)) -> df_ind

print(df_ind)

# TODO David: Zahlen überprüfen
write.csv(file = "output/datasets/202207_table_kiefer_individual_hypodontia.csv", df_ind)


#double check it: seems to fit!
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>%
  filter(!duplicated(ID)) %>%
  #change some cases: o, o to o; u,u to u, o, u to beide
  mutate(oben_unten = str_replace(oben_unten, "o, o", "o")) %>%
  mutate(oben_unten = str_replace(oben_unten, "u, u", "u")) %>% 
  mutate(oben_unten = str_replace(oben_unten, "o, u", "beide")) %>%
  # filter(!is.na(oben_unten)) %>% 
  group_by(oben_unten, gschlecht_m_w) %>% 
  summarise(total = n()) %>%
  mutate(pct = total / sum(total))-> ttt



#table links - rechts-----------
# df_t %>% 
#   mutate(quadrant = stringr::str_sub(fehlender_zahn_fdi_quadrant_zahnzahl, end = 1)) %>% 
#   mutate(quadrant = as.numeric(quadrant)) %>% 
#   group_by(ID) %>% 
#   summarise(tot = sum(quadrant)) %>% 
#   ungroup() %>% 
#   mutate(tot = if_else(tot == 3, 2.5, tot)) %>% #weil summe beider seiten 5 ergeben
#   mutate(seite = case_when((tot == 5 | tot == 1 | tot == 4) ~ "rechts",
#                             (tot == 4.5 | tot == 2 | tot == 2.5) ~ "links",
#                             TRUE ~ "beide")) ->  df_seite 
# 
# #some cases need to be changes manually
# df_seite[df_seite$ID == 67, ]$seite <- "beide" #sum = 3
# df_seite[df_seite$ID == 185, ]$seite <- "beide" #sum = 3
# df_seite[df_seite$ID == 243, ]$seite <- "beide" #sum = 3
# df_seite[df_seite$ID == 419, ]$seite <- "beide" #sum = 3
# df_seite[df_seite$ID == 472, ]$seite <- "beide" #sum = 4
# df_seite[df_seite$ID == 480, ]$seite <- "beide" #sum = 3
# df_seite[df_seite$ID == 508, ]$seite <- "beide" #sum = 3


# #group again
# df_seite %>% 
#   group_by(seite) %>% 
#   summarise(tot = n()) %>% 
#   mutate(pct = tot / sum(tot)) -> df
# 
# print(df)


#1. individual level: prepare data
data %>% 
  #age older than 9
  filter(alter_beim_opt >= 9) %>%
  #drop all cases with etrahierte zähne
  filter(is.na(extrahierte_zahne)) %>% 
  filter(fehlender_zahn_fdi_quadrant_zahnzahl > 0) %>%
  #drop dupls
  filter(!duplicated(ID)) %>%
  #change some cases
  mutate(links_rechts = str_replace(links_rechts, "l, l", "l")) %>%
  mutate(links_rechts = str_replace(links_rechts, "r, l", "beide")) %>%
  select(ID, gschlecht_m_w, links_rechts, fehlender_zahn_fdi_quadrant_zahnzahl) -> df


#group data
df %>%
  group_by(links_rechts, gschlecht_m_w) %>% 
  summarise(total = n()) %>%
  mutate(pct = total / sum(total)) -> df_ind


# TODO David: Zahlen überprüfen
write.csv(file = "output/datasets/202207_table_seite_individual_hypodontia.csv", df_ind)

#prepare data for chi_square
xtabs(~ links_rechts + gschlecht_m_w, data = df) -> df_chisq

chisq.test(df_chisq) # nicht signifikant



```

